---
title: "Latent Transition Analysis"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, anthroheightcalc, anthroweightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc, anthroweightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2", "Year_3")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2),
        anthroweightcalc = round(as.numeric(anthroweightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = anthroheightcalc,
        weight = anthroweightcalc
    )

df_long <- df_long %>%
    group_by(id) %>%
    filter(all(c("Baseline", "Year_1", "Year_2", "Year_3") %in% event)) %>%
    ungroup()

```

## Overview

XXXXXXXXXXX Models xxxxxxx. This xxxxxxx.

In this example, we will utilize xxxxxx models to analyze xxxxxx obtained across multiple measurement occasions for a sample of youth participating in the ABCD Study. Our primary objective is to understand xxxxxx, while factoring in the clustered nature of observations within individuals over time. XXXX models facilitate this by xxxxx.

## Preliminary Setup
### Load R libraries
```{r}
library(tidyverse)    # Collection of R packages for data science
library(rstatix)      # Pipe-friendly framework for basic statistical tests
library(DT)           # Rendering interactive data tables
library(tidyLPA)      # XXXXXXXX

```

### Descriptives
```{r}
descriptives_table <- df_long %>%
    select(event, height, weight) %>%
    mutate(event = factor(event)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(height ~ "Height", weight ~ "Weight"),
        statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table
```

## Results 
### Compute Latent Transition Analysis
This code performs latent transition analysis to examine the reciprocal relationships between 'Height' and 'Weight' measurements across 3 annual visits. The model output is provided below, along with a summary of results. This approach xxxxx.

```{r}
# Estimate the LTA model
model <- tidyLPA::estimate_profiles(df_wide, 2, variances = "equal", covariances = "zero")

summary(model, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)

plot(model)

```

These results show xxxxx, indicating that xxxxx (e.g., xxxxx). The xxxx effects indicate xxxx.

### Model Plots
```{r}

```

This plot illustrates xxxx.

## Wrapping Up
The multivariate latent growth curve revealed xxxxxxxx.

