---
title: "midFEMA Manuscript"
authors:
  - name: Samuel W. Hawes
    email: "shawes@fiu.edu"
    affiliations:
      - name: Florida International University
        department: Psychology
        city: Miami
        state: Florida
        country: United States
  - name: Mohammadreza Bayat
    email: "mbayat@fiu.edu"
    affiliations:
      - name: Florida International University
        department: Psychology
        city: Miami
        state: Florida
        country: United States
  - name: Anthony Dick
    email: "adick@fiu.edu"
    affiliations:
      - name: Florida International University
        department: Psychology
        city: Miami
        state: Florida
        country: United States
  - name: Raul Gonzalez
    email: "gonzara@fiu.edu"
    affiliations:
      - name: Florida International University
        department: Psychology
        city: Miami
        state: Florida
        country: United States
  - name: Additional Co-authors
    email: "tbd@fiu.edu"
    affiliations:
      - name: Florida International University
        department: Psychology
        city: Miami
        state: Florida
        country: United States

abstract: |
  The midFEMA research project is aimed at identifying and analyzing neurodevelopmental biomarkers for severe behavioral and emotional problems in youth. By leveraging Fast and Efficient Mixed‐Effects Analysis (FEMA) in conjunction with voxelwise task-based reward activation measurements, we aim to uncover insights into brain development and functioning across a large, prospectively followed sample of approximately 9,000 youth aged 9–16 years from the Adolescent Brain and Cognitive Development (ABCD) study.
index-terms: 
  - Longitudinal
  - fMRI
  - DBD
  - FEMA
  - MID
format:
  ieee-typst: default
bibliography: references.bib
---

```{r}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(lorem)
```

```{r}
library(tidyverse)
library(ggplot2)

df <- mtcars
str(df)
```


```{r}
# Set the URL for the CSV file
file_url <- "https://drive.google.com/uc?export=download&id=1QFdAFz09DQEeFVOj8bRWh4g6hK4b4Qj1"

# Set the destination filename
dest_file <- "df_preclean.csv"

# Download the file
download.file(file_url, destfile = dest_file, mode = "wb")

# Read the CSV file into R
my_data <- read.csv(dest_file)

# Display the first few rows of the data
head(my_data)

```


## Introduction
_Introduce the research topic, state the research problem, and outline the study's objectives. Provide background information to set the context. This is an exmaple in-text citation_ @netwok2020.
`r lorem::ipsum(paragraphs = 3)`


### Background
_Provide a brief review of the pertinent literature, highlighting the gaps your study aims to fill._
`r lorem::ipsum(paragraphs = 2)`


### Objectives
_State the specific research questions or hypotheses your study addresses._
`r lorem::ipsum(paragraphs = 2)`


## Methods
_Describe the methodological approach, detailing how the study was conducted, how data was collected, and how analyses were performed._
`r lorem::ipsum(paragraphs = 1)`


### Participants
_Describe the demographic characteristics of the study participants, including how they were selected, recruited, and any inclusion or exclusion criteria._
`r lorem::ipsum(paragraphs = 1)`


### Measures
_List the instruments or tools used to collect data, describing each measure's purpose and any relevant psychometric properties._
`r lorem::ipsum(paragraphs = 2)`


### Analytic Plan
_Detail the statistical tests or models employed to analyze the data, including any software used._
`r lorem::ipsum(paragraphs = 1)`


## Results
Present the findings of the study, supported by statistical analysis, tables, and figures as necessary.
`r lorem::ipsum(paragraphs = 3)`


### [Subsection Title]
_Present specific findings related to a particular aspect of the study, supported by data._
`r lorem::ipsum(paragraphs = 1)`


### [Subsection Title]
_Discuss another set of findings, providing details on the analysis and interpretation._
`r lorem::ipsum(paragraphs = 1)`


## Discussion
_Interpret the results, discussing their implications in the context of existing literature. Address the study's limitations and suggest directions for future research._
`r lorem::ipsum(paragraphs = 3)`


### Limitations
_Acknowledge any limitations of your study and their potential impact on the findings._
`r lorem::ipsum(paragraphs = 1)`


### Future Directions
_Suggest areas for future research that could build on your findings or address limitations._
`r lorem::ipsum(paragraphs = 1)`


## Conclusion
_Summarize the main findings and their implications. Conclude with a statement that reflects the significance of the study and its contributions to the field._
`r lorem::ipsum(paragraphs = 1)`


## Acknowledgments
_Acknowledge anyone who contributed to the research but does not meet authorship criteria, along with any funding sources._
`r lorem::ipsum(paragraphs = 1)`


## Appendices
_Include any additional material that is pertinent to the study but would be too distracting or bulky to include in the main body of the manuscript._
`r lorem::ipsum(paragraphs = 1)`



