---
title: "Difference Scores: Simple Regression"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, ehi_y_ss_scoreb, anthroheightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    group_by(src_subject_id) %>%
    fill(ehi_y_ss_scoreb, .direction = "downup") %>%
    ungroup() %>%
    filter(ehi_y_ss_scoreb %in% c(1, 2)) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1")),
        ehi_y_ss_scoreb = factor(ehi_y_ss_scoreb, levels = c(1, 2), labels = c("Right", "Left")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        handedness = ehi_y_ss_scoreb,
        height = anthroheightcalc
    )

```

## Overview
Difference scores can be used to assess change in a variable across two timepoints. The value of the variable at the first timepoint is substracted from the value at the second timepoint. A simple regression can then be applied to determine whether a variable of interest predicts the computed difference score.

This example will examine height values obtained across 2 annual measurement occasions in a sample of youth taking part in the ABCD Study. The primary aim is to determine whether a variable of interest (handedness) predicts the difference in height values between the two timepoints. A difference score will be calculated between participant's height values collected at baseline and again at the 1-year follow-up. Next, a simple regression analysis will be conducted to test whether handedness (left, right) predicts the average difference value in participants' height from baseline to the 1-Year follow-up. Finally, a visual inspection is further conducted via a violin plot to graphically represent the relationship between the difference score and handedness.

## Preliminary Setup
### Load R libraries
```{r}

library(tidyverse) # Collection of R packages for data science
library(gtsummary) # Publication-ready tables
library(rstatix) # Statistical Tests in tidy format

```

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, handedness, height) %>%
    mutate(event = factor(event)) %>%
    mutate(handedness = factor(handedness)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(handedness ~ "Handedness", height ~ "Height"),
        statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

```

## Results
### Compute Difference Score
This code computes a difference score by subtracting participant's height at baseline from height at the 1-year follow-up. Next, a simple regression analysis is conducted to determine whether handedness predicts the difference in height values between the two timepoints.

```{r}

# Reshape data from long to wide format
df_wide <- df_long %>%
  pivot_wider(
    id_cols = c(id, handedness),
    names_from = event,
    values_from = height,
    names_prefix = "Height_"
  )

# Compute the height difference score
df_wide <- df_wide %>%
    mutate(
        height_diff = Height_Year_1 - Height_Baseline
    )

# Calculate summary statistics for the difference score
summary(df_wide$height_diff)

# Compute summary statistics for Height by eventname
summary <- df_long %>%
    group_by(event) %>%
    get_summary_stats(height, type = "mean_sd")

data.frame(summary)

```
The summary statistics provide insights into participant's height across two annual assessments. At baseline, the average height is 55.24 inches (sd = 3.33). Over the year, there is an increase of 2.35 inches in average height to 57.59 inches at the time of the 1-year follow-up.

### Conduct regression on Difference Score
A simple regression analyses is conducted to examine whether a grouping variable (handedness) predicts the difference score value.
```{r}

# Run the regression
model <- lm(height_diff ~ handedness, data = df_wide)

# Get the summary of the regression model
summary(model)

```

The results from the simple regression analysis suggest handedness does not significantly predict change in height. The intercept, estimated at 2.35 with a standard error of .02, indicates the average height difference for right-handed individuals is approximately 2.35 units, with this effect differing by only .03 inches (p=.69) for left-handed individuals

### Model Plots
```{r}

# Visualize the difference scores across different levels of handedness

# Create a violin plot to show the distribution of difference scores by sex
# Jittered points are added to provide a more granular view of individual observations
# Load necessary libraries
library(RColorBrewer)

# Plotting using ggplot2
ggplot(df_long, aes(x = handedness, y = height, fill = handedness)) +
    geom_violin() +
    geom_jitter(position = position_jitter(width = 0.2), size = 1, alpha = 0.5) +
    scale_fill_brewer(palette = "Set2") +
    labs(
        title = "Difference Scores by Handedness",
        x = "Handedness",
        y = "Height Difference Score"
    ) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none"
    )

```

The violin plot visualizes the distribution of difference scores in children's heights by their handedness (Right, Left). Each violin shape provides insight into the density of the data at different height differences, with wider sections representing higher densities of data points. Superimposed jittered points offer a granular view of individual observations. From the plot, it appears that the distributions of height differences across the two handedness categories are markedly similar.

## Wrapping Up
This analysis examined difference scores of participant's height across a one-year period and used a simple regression analysis to explore the impact of handedness on the year-to-year change in children's height. Findings showed an increase in height values from baseline (55.23 inches) to the year 1 follow-up (57.57 inches). The regression model's output shows that handedness, specifically being left-handed as compared to right-handed, is not associated with any change in height (r=.03, se=.08, p=.69). Further, a violin plot visualizing the distribution of difference scores in children's heights by their handedness showed no substantive effect. 

This difference score approach used in this example quantifies the change in height values over time and uses a simple regression to test whether participant handedness predicts this value. More generally, this approach is often implemented when data used to evalaute change is only available at two timepoints, as more flexible approaches are typically implemented when data from additional (>2) time points is available.
