---
title: "Add Example Title Here"
author: "Name Here"
---

## Overview

<p>

[ Insert a brief description of the analysis and its purpose here. ]

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

# Create a list of required packages
packages_required <- c("addpackage", "addpackage", "addpackage")

# Check which packages are not installed and install them
packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
if (length(packages_to_install) > 0) {
    install.packages(packages_to_install)
}

# Load the required packages
lapply(packages_required, library, character.only = TRUE)

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(addpackage) # snippet about the package
library(addpackage) # snippet about the package
library(addpackage) # snippet about the package

```
:::

### Config Options

::: blue
**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```
:::
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue

**This code reads in and shows the data to be used in the current example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

# Read and wrangle data

```
:::

### Descriptives

::: blue
**This code creates a descriptives table**

```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE


```

:::

## Results

::: blue
[Brief explanation of the code]
```{r build-model-ttest-1}
#| message: FALSE
#| warning: FALSE
#| echo: TRUE


```

:::
:::

:::

### Model Summary {.tabset .tabset-fade .tabset-pills}

::: blue

A plot to show xxxxx.
```{r model-checking}

```

[brief write-up/explanation of the results]

:::

### Model Plots  {.tabset .tabset-fade .tabset-pills}

::: blue
```{r scatterplot}


```

:::
:::

## Wrapping Up

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue

[insert brief write-up here]
  

:::
:::
